EDA Project 2 -- Air Pollution Dataset Workbook
===============================================

Data Loading
------------

```{r pollutant data loading, cache = TRUE}
setwd('~/airpol')
download.file('https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip',method='curl',destfile='NEI_data.zip')
unzip('NEI_data.zip')
NEI <- readRDS("summarySCC_PM25.rds")
NEI <- transform(NEI,fips=factor(fips),SCC=factor(SCC),Pollutant=as.factor(Pollutant),type=as.factor(type),year=as.factor(year))
SCC <- readRDS("Source_Classification_Code.rds")
```
Data Checking
-------------

Make sure I've got what I want ...

```{r Pollutant types check, cache = TRUE}
foo <- as.factor(NEI$Pollutant)
levels(foo)
rm(foo)
```

Question 1
----------

__Question:__ Have total emissions from PM2.5 decreased in the United States from 1999 to 2008?
Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the
years 1999, 2002, 2005, and 2008.

I'm gonna do this in __Plot1.R__

```{r Get year totals, cache = TRUE}
yearTotals <- as.data.frame(rowsum (NEI$Emissions, NEI$year, reorder = TRUE))
```

```{r Year total barplot, fig.width=5, fig.height=5}
barplot (yearTotals$V1, names.arg = row.names(yearTotals), main = "Nationwide Annual PM25 Pollution Totals" )
```

Question 2
----------

__Question:__ Have total emissions from PM2.5 decreased in the Baltimore City, Maryland
(fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot
answering this question.

Done in __Plot2.R__

```{r Filter down to Baltimore, cache = TRUE}
baltiNEI <- NEI[NEI$fips=='24510',]
baltiYearTotals <- as.data.frame(rowsum (baltiNEI$Emissions, baltiNEI$year, reorder = TRUE))
```
```{r Plot Baltimore Year Totals, fig.width=5, fig.height=5}
barplot (baltiYearTotals$V1, names.arg = row.names(baltiYearTotals), main = "Baltimore Annual PM25 Pollution Totals" )
```

Question 3
----------

__Question:__ Of the four types of sources indicated by the type (point,
nonpoint, onroad, nonroad) variable, which of these four sources have seen
decreases in emissions from 1999–2008 for Baltimore City? Which have seen
increases in emissions from 1999–2008? Use the ggplot2 plotting system to make
a plot answer[ing] this question.

Done in __Plot3.R__

```{r recast Baltimore year/type data}
library ("reshape2")
baltiMelt <- melt (baltiNEI, id.vars = c("type", "year", "fips", "SCC", "Pollutant"), value.name = "Emissions")
baltiYearType <- dcast (baltiMelt, type+year ~ variable, fun.aggregate = sum)
```
```{r Baltimore year/type exploratory plot, fig.width = 8, fig.height = 5}
g<-ggplot(baltiYearType,aes(year,Emissions))
g+geom_bar(stat='identity') + facet_grid( . ~ type )
```

Question 4
----------

__Question:__ Across the United States, how have emissions from coal
combustion-related sources changed from 1999–2008?

Done in __Plot4.R__

```{r Analysis of nationwide coal combustion sources}
coalTypes <- SCC [grepl ('coal', SCC$Short.Name, ignore.case = TRUE) & grepl ('comb',S CC$Short.Name, ignore.case =TRUE), ]
coalCombNEI <- NEI [is.element (NEI$SCC, coalTypes$SCC), ]
coalCombYearTotals <- as.data.frame (rowsum (coalCombNEI$Emissions, coalCombNEI$year, reorder = TRUE))
```
```{r Nationwide coal combustion plot, fig.width = 5, fig.height = 5}
barplot (coalCombYearTotals$V1, names.arg = row.names (coalCombYearTotals), main = "Nationwide Annual Coal Combustion PM25 Totals" )
```

Question 5
----------

__Question:__ How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?

Done in __Plot5.R__

Actually quite difficult to boil it down to cars, motorbikes, trucks, what else? Construction vehicles maybe?

```{r filter motor vehicle sources -- first try}
possibleVehTypes <- SCC[
(!grepl('ext comb|int comb|stationary|locomotive|coating|aircraft|pleasure craft|commercial equipt',SCC$Short.Name,ignore.case=TRUE))&
(!grepl('agriculture|fuel comb|cooking|^dust|fires|^industrial|solvent|locomotives|marine|aircraft|waste disposal|NEC|gas stations|non-road equipment',SCC$EI.Sector,ignore.case=TRUE))
,]
```

... ah, fuhgeddaboutit ...

```{r filter motor vehicle sources -- second try}
vehTypes <- SCC[grepl('veh',SCC$EI.Sector,ignore.case=TRUE),]
baltiVeh <- baltiNEI <- NEI[NEI$fips=='24510'&is.element(NEI$SCC,vehTypes$SCC),]
baltiVehYearTotals <- as.data.frame(rowsum (baltiVeh$Emissions, baltiVeh$year, reorder = TRUE))
```
```{r Baltimore vehicles emissions plot, fig.width = 5, fig.height = 5 }
barplot (baltiVehYearTotals$V1, names.arg = row.names(baltiVehYearTotals), main = "Baltimore Annual Vehicle PM25 Pollution" )
```

Question 6
----------

__Question:__ Compare emissions from motor vehicle sources in Baltimore City
with emissions from motor vehicle sources in Los Angeles County, California
(fips == "06037"). Which city has seen greater changes over time in motor
vehicle emissions?

Done in __Plot6.R__

```{r Get Baltimore + LA Vehicle Data}
baltiLosAngVeh <- NEI[(NEI$fips=='24510'|NEI$fips=='06037')&is.element(NEI$SCC,vehTypes$SCC),]
library('reshape2')
baltiLaMelt <- melt (baltiLosAngVeh, id.vars = c("type", "year", "fips", "SCC", "Pollutant"), value.name = "Emissions")
fipsYear <- dcast (baltiLaMelt, fips+year ~ variable, fun.aggregate = sum)
```
```{r plot Baltimore + LA Vehicles, fig.width = 8, fig.height = 5 }
library('ggplot2')
g<-ggplot(fipsYear,aes(year,Emissions))
g + geom_bar(stat='identity') + facet_grid( . ~ fips ) + scale_y_log10(breaks = c(10,100,1000)) + ylab('Emissions -- Log Scale') + ggtitle("LA & Baltimore Vehicle Emissions")
```
